---
title: |-
  R Crash Course \
   Part 1 -- Fundamentals TEST
author: "Rob Colautti"
---

<script src="_hidecode.js"></script>

# Test your understanding of R Fundamentals

<br>

***

<br>

# 2. R Basics

Make comments inside your code. Very important (unless you are using R markdown or R notebooks)!
<div class="fold s o">
```{r}
# Use hastags to make comments - not read by the R console
# Use other characters and blank lines to improve readability:
# ------------------------- 
# My first R script 
# Today's Date
# -------------------------
# Add a summary description of what the script does
# This script will...
# And annotate individual parts of the script
```
</div>

## Basic Math

add, subtract, multiply, divide, exponent
<div class="fold s o">
```{r, eval=F}
10+2 # add
10-2 # subtract
10*2 # multiply
10/2 # divide
10^2 # exponent
```
</div>

absolute value, square root, natural log (ln), log base 10, e to the power of 6
<div class="fold s o">
```{r, eval=F}
abs(-10) # absolute value
sqrt(10-1) # square root (with subtraction)
log(10) # natural log
log10(10) # log base 10
exp(6) # power of e
```
</div>


sine, cosine, tangent and their inverse
<div class="fold s o">
```{r, eval=F}
sin(pi/2) # sine function
asin(1) # inverse sine
cos(pi) # cosine
acos(-1) # inverse cosine
tan(0) # tangent
atan(0) # inverse tangent
```
</div>

## Round/Truncate
round to 3 decimal places; round down; round up; keep 2 significant digits
<div class="fold s o">
```{r, eval=F}
round(pi,digits=3) # standard rounding to 3 digits
floor(pi) # round down to closest whole number
ceiling(pi) # round up to closest whole number
signif(pi,digits=2) # round to keep 2 significant digits
```
</div>

## Logic Operators

greater, less, equal, not equal, 'OR' 'AND'
<div class="fold s o">
```{r, eval=F}
1 > 2 # greater than
1 < 2 # less than
1 <= 2 # less than or equal to
1 == 1 # equal to
1 != 1 # not equal to
1 == 2 | 1 # | means 'OR'
1 == 2 & 1 # & means 'AND' 
```
</div>

## Random Number Generators

Generate 10 numbers from each of the following:

Uniform distribution
<div class="fold s">
```{r}
runif(10,min=0,max=1) 
```
</div>

Normal distribution with a mean of 10 and standard deviation of 5
<div class="fold s">
```{r}
rnorm(10,mean=0,sd=1)
```
</div>

Poisson distribution with a lambda of 100
<div class="fold s">
```{r}
rpois(10,lambda=10)
```
</div>

Bernoulli variable (Binomial distribution) to model a coin toss (one coin, 10 separate trials)
<div class="fold s">
```{r}
rbinom(10,size=1,prob=0.5)
```
</div>

Bernoulli variable (Binomial distribution) to model the number of Adenine nucleotides found in a stretch of 100bp of DNA (assuming equal representation of the 4 base pairs) -- repeated for 10 independent strands of DNA
<div class="fold s">
```{r}
rbinom(10,size=100,prob=0.25)
```
</div>

## Combining objects

Make a vector of 3 letters
<div class="fold s">
```{r}
c('a','b','c')
```
</div>

Make a vector of numbers 1 to 100
<div class="fold s">
```{r}
c(1:100)
```
</div>

Make a vector of numbers counting from 1.1 to 10.1 by 0.1
<div class="fold s">
```{r}
c(11:101)/10
```
</div>

<br>

*** 

<br>

# 3. Defining variables / objects

## Cells

Define an object $X$ as the word $string$
<div class="fold s">
```{r}
X<-"string"
```
</div>

A few options To see the contents of **X**:
<div class="fold s">
```{r, eval=F}
# Any of these will work:
print(X)
paste(X)
X
```
</div>

#### What's the difference?

What is the difference between `X`, `print(X)` and `paste(X)`? 
<div class="fold s">
```{r eval=F}
# print() Is most generic, useful for providing feedback from running scripts (e.g. during loops, Bash scripts, etc)

# paste() Converts objects to a string, we'll come back to this.

# Generally print() or paste() are preferred over calling the object directly.
```
</div>

### Subsetting objects 

Make a vector of numbers 1 to 3.6 by 0.1 called `Xvec` 

<div class="fold s">
```{r}
Xvec<-c(1:36)*0.1
```
</div>

Print the 1st, 3rd and 9th elements of `Xvec` as a single string

<div class="fold s">
```{r}
Xvec[c(1,3,9)]
```
</div>

## Matrices 

Make a matrix called `Xmat` by converting the 1x36 `Xvec` vector to a 6x6 matrix
<div class="fold s">
```{r}
Xmat<-matrix(Xvec,nrow=6)
```
</div>

### Subset a matrix

Print the number from the 1st row, 3rd column of the matrix object `Xmat`
<div class="fold s">
```{r}
Xmat[1,3]
```
</div>

## Higher-order arrays

Make an object containing 2 3x3 matrices (i.e. 3-dimensional array)
<div class="fold s o">
```{r}
Xarray<-array(0, dim=c(3,3,2)) # 3 dimensions
Xarray
```
</div>

### Matrix Algebra Basics

Start with a couple of vectors
```{r}
X<-c(1:10)
Y<-c(10:1)
```

Multiply each element of the vector X by its corresponding element in vector Y
i.e. `X[1]` x `Y[1]`, `X[2]` x `Y[2]`, etc.
<div class="fold s">
```{r}
X*Y 
```
</div>

Take every element of Y and raise it to the exponent $e^Y$
<div class="fold s">
```{r}
exp(Y)
## Remember: you can use pretty much any standard operator for element-by element calculations
```
</div>

## More advanced matrix algebra

Standard ('proper') matrix multiplication of vector `X` by vector `Y`. 

i.e. `X[1]*Y[1]+X[2]*Y[2]`

What will the output be?

<div class="fold s o">
```{r}
X%*%Y
```
</div>

Outer product of `X` and `Y`.

What will the output be?
<div class="fold s o">
```{r}
Z<-X%o%Y
Z
```
</div>

### Principal components analysis

Do a principal components analysis of the matrix `Z` from the previous step
<div class="fold s o">
```{r}
prcomp(Z) # or princomp(Z)
```
</div>

## Lists 

Make a list called _MyList_ containing the following elements:

1. A string called _name_ that contains your name  
2. A vector called _Xvec_ that all the elements of the vector X, defined in section 3  
3. A vector called _numbers_ that contains 10 random numbers

What will the output look like?

<div class="fold s o">
```{r}
MyList<-list(name="Rob",Xvec=X,numbers=runif(10)) # if your name is 'Rob'
MyList
```
</div>

Print the _numbers_ element from the MyList list that you created
<div class="fold s">
```{r}
MyList$numbers # Use $ to subset by name
```
</div>
<br>

What's the difference between `$`, `[]` and `[[]]`?

What will the ouput look like (predict all 6 before revealing)?
<div class="fold o">
```{r, error=TRUE}
MyList$numbers
```
</div>

<div class="fold o">
```{r, error=TRUE}
MyList[3]
2*MyList[3]
```
</div>

<div class="fold o">
```{r, error=TRUE}
MyList[[3]]
2*MyList[[3]]
```
</div>

Describe the above in words:
<div class="fold s">
```{r, eval=F}
# The [] his the whole 'slice' including the title $numbers
# The [[]] includes only the values, similar to $
```
</div>

### Protip: 

Don't forget: many analysis functions in R output as lists (e.g. statistical packages)

<br>

For example, the output of prcomp is a list that can be sliced:
```{r}
prcomp(Z) 
names(prcomp(Z))
prcomp(Z)$center
prcomp(Z)$scale
```

## paste() function

A very versatile function for manipulating strings, including output from other functions

Paste two strings _Hello_ and _World!_
<div class="fold s">
```{r}
paste("Hello","World!")
```
</div>

Paste a vector of numbers c(1:10) as strings
<div class="fold s">
```{r}
paste(c(1:10))
```
</div>

Paste a vector of numbers c(1:10) as a single string, separated by a comma and a space
<div class="fold s">
```{r}
paste(c(1:10),collapse=". ")
```
</div>

# 4. Working with data

## Working Directory
Tell R which directory you want to work from
<div class="fold s">
```{r, eval=F}
setwd("C:/Users/ColauttiLab/Documents")
```
</div>

Ask R which directory it is currently working in
<div class="fold s">
```{r, eval=F}
getwd()
```
</div>

## Import data

Import data from _FallopiaData.csv_ file into an object called _MyData_
<div class="fold s">
```{r}
MyData<-read.csv("FallopiaData.csv",header=T)
```
</div>

## **`data.frame`**

What is a **`data.frame`** object, how is it different from other objects?

<div class="fold s">
```{r}
# A data.frame is a special type of 2D matrix with additional indexing information for rows/columns of data
```
</div>

Inspecting your **`data.frame`** object:

Print the names of the columns in your `data.frame` called _MyData_
<div class="fold s">
```{r}
names(MyData) # See column names
```
</div>

Inspect the first few rows of _MyData_
<div class="fold s">
```{r}
head(MyData) #  Show first six rows of data
```
</div>

Inspect the last few rows
<div class="fold s">
```{r}
tail(MyData) #  Show last six rows of data
```
</div>

Find the number of rows and columns of _MyData_  
<div class="fold s">
```{r}
dim(MyData) #  Number of rows x columns (or 'dimension') of the data object; nrow(MyData) and ncol(MyData) also would work
```
</div>

Figure out what class of variables you have for each column of _MyData_  

<div class="fold s">
```{r}
str(MyData) #  Data 'structure' - types of variables
```
</div>

Why is it important to check the class (e.g. integer, numeric, string, factor) of your `data.frame` object?

<div class="fold s">
```{r}
# str() is very important for functions that use data.frames including statistical analysis and plotting. For example if you have two treatment groups coded as 1 and 2, this should be a FACTOR, but R will default to an INTEGER. This could be a problem if your analysis treats INTEGERS as continusous variables.

# Another common problem is when you have a typo in a measurement column (e.g. you accidentally press o instead of 0 for one of your measurements). Since there is one letter in the column, R will treat it as a string or factor, rather than a continuous measurement.
```

How would you convert the __PotNum__ column in __MyData__ from an integer to a factor?
<div class="fold s">
```{r eval=F}
# Just replace the column
MyData$PotNum<-as.factor(MyData$PotNum)
```
</div>


## Adding calculations

Add a new column in __MyData__ called __Total__ that is the sum of __Symphtum__, __Silene__, __Urtica__, __Geranium__, __Geum__ and __Fallopia__
<div class="fold s">
```{r eval=F}
MyData$Total<-MyData$Symphytum+MyData$Silene+MyData$Urtica+MyData$Geranium+MyData$Geum+MyData$Fallopia
```
</div>

<br>

*** 

<br>

# 5. Summary statistics

Working with the __MyData__ **`data.frame`** 

Calculate means of the __Total__ column you created above, but for each level of __Nutrients__ column
<div class="fold s">
```{r}
aggregate(Total~Nutrients,data=MyData,mean)
```
</div>

Calculate standard deviations instead of means
<div class="fold s">
```{r}
aggregate(Total~Nutrients,data=MyData,sd) 
```
</div>

Calculate means but separately for each combination of factors in the __Nutrients__, __Taxon__ and __Scenario__ columns
<div class="fold s">
```{r}
aggregate(Total~Nutrients*Taxon*Scenario,data=MyData,mean)
```
</div>

Alternative answers to the above (note difference in output)
<div class="fold s o">
```{r}
tapply(MyData$Total,list(MyData$Nutrients),mean) # calculate means
tapply(MyData$Total,list(MyData$Nutrients),sd) # calculate standard deviation
```
</div>

<br>

*** 

<br>

# 6. Save output


Create new object containing the means of __Total__ for each level of __Nutrients__

Then save it as a csv file called "**MyData_Nutrients_Means.csv**"

<div class="fold s">
```{r}
## Calulate means
NutrientMeans<-tapply(MyData$Total,list(MyData$Nutrients),mean)
## Save means as .csv file
write.csv(NutrientMeans,"MyData_Nutrient_Means.csv")
```
</div>

<br>

*** 

<br>

# 7. Flow control

## Brief exampes

Make up a couple of ojects to play with
```{r}
X<-21
Xvec<-c(1:10,"string")
```

### **`if(){}`**

Asses the object __X__ and if it is greater than 100, tell the user "X > 100", but if it not, then tell the use "X <= 100"

<div class="fold s">
```{r, eval=F}
if(X > 100){
  print("X > 100")
} else {
  print("X <= 100")
}
```
</div>

### **`for(){}`**

Create a loop that will repeat __X__ times, and in each loop, tell the user which iteration it is on.
<div class="fold s">
```{r}
# Loop through numbers from 1 to X
for (i in 1:X){
  print(paste(i,X,sep=" of "))
}
```
</div>

Create a loop that will read each element of Xvec and tell the user which element is being read.
<div class="fold s">
```{r}
# Loop through elements of a vector directly
for (i in Xvec){
  print(i)
}
```
</div>

Create a nested loop that will first cycle through the first 3 columns of __MyData__ and then for each column, cycle through the the first 4 rows. Keep track of the total number of iterations. For each iteration, tell the user:

1. What iteration it is on: 'Iteration X of Y'
2. What row of __MyData__ is being read
3. What column is being read
4. The element (value) in that particular row x col
<div class="fold s">
```{r}
# Use an index to loop through the elements
iter<-1
for (i in 1:3){
  for (j in 1:4){
  print(paste("Iteration:",iter,"of",3*4))
  print(paste("Element =",MyData[j,i],"at row:",j,"and col:",i))
  iter<-iter+1
  }
}
```
</div>

### **`while(){}`**

Randomly sample values for 2 variables, each drawn from a random normal distribution with mean = 0 and sd = 1.
Repeat until the product is a value greater than 1, and output the number of iterations it took.

```{r,echo=F}
set.seed(2)
```

<div class="fold s">
```{r}
count<-0
while (rnorm(1)*rnorm(1)<=1){
  count<-count+1
}
print(paste("It took",count,"iterations to reach a value > 1"))
```
</div>