---
title: |-
  R Crash Course \
   Part 3 -- Customizing *ggplot()*
author: "Rob Colautti"
---

# Getting Started
Install the `ggplot2` package

```{r}
library(ggplot2)
```


# 1. Setup

## ppt slides
  1. Graphical concepts

  2. ggplot grammar
  
  3. Explanation of selection meta-analysis data

  4. anatomy of a graph
  
  [Slides](Graphics_small.pdf)
  

<br>

## Import data

Download selection dataset from Colautti & Lau (2015) -- this data is archived on Dryad:

http://datadryad.org/

```{r}
SelData<-read.csv("http://datadryad.org/bitstream/handle/10255/dryad.82839/Selection_Data.csv?sequence=1",header=T)
```

#### Change column names 
To make them more intuitive in R
```{r}
names(SelData)<-c("Collector","Author","Year","Journal","Vol","Species","Native","N","Fitness.measure","Trait","s","s.SE","s.P","B","B.SE","B.P")
```

#### Replace s with its absolute value 
We are interested in magnitude, not direction in the meta-analysis
```{r}
SelData$s<-abs(SelData$s)
```

#### Add random variables
We'll use these later to explore some additionall ggplot options
```{r}
SelData$Rpoint<-rnorm(nrow(SelData)) # Random, normally distributed
SelData$Rgroup<-sample(c(0,1),nrow(SelData),replace=T) # Random binary value
```

#### A quick look at the data
```{r}
head(SelData)
```

#### One more thing...
Note the missing data (denoted NA)
```{r, eval=FALSE}
print(SelData$s)
```

We can subset to remove mising data
```{r}
SelData<-SelData[!is.na(SelData$s),]
```

<br>

***

<br>

# 2. *ggplot()*  vs  *qplot()*

## Histogram
#### qplot
```{r, error=TRUE}
BarPlot<-qplot(s, data=SelData, fill=Native, geom="bar")
print(BarPlot)
```

#### ggplot
```{r, error=TRUE}
BarPlot <-ggplot(aes(s, fill=Native), data=SelData) 
print(BarPlot)
```
No layers! We only loaded in the data info for plotting

We have to specify which geom(s) we want
```{r}
BarPlot<- BarPlot + geom_bar() # info from ggplot() passed to geom_bar()
BarPlot
```

Explore the components of our BarPlot object:
```{r}
summary(BarPlot)
```

For more information on geom_bar()
```{r, eval=FALSE}
?geom_bar
```

### Bivariate geom
```{r}
BivPlot<-ggplot(data=SelData, aes(x=s, y=Rpoint)) + geom_point()
print(BivPlot)
```

Add linear regression
```{r}
BivPlot + geom_smooth(method="lm",colour="steelblue",size=2)
```

Add separate regression lines for each group
```{r}
BivPlot + geom_smooth(method="lm",size=2,aes(group=Native,colour=Native))
```

<br>

***

<br>

# 3. Full ggplot
Recreate the selection histograms from Colautti & Lau:

  1. Create separate data for native vs. introduced species

  2. Use a bootstrap to estimate non-parametric mean and 95% confidence intervals
  
  3. Plot all of the components on a single graph

## 1. Separate data
```{r}
NatSVals<-SelData$s[SelData$Native=="yes"] # s values for Native species
IntSVals<-SelData$s[SelData$Native=="no"] # s values for Introduced species
```

## 2. Bootstrap
### 2a. Setup
```{r}
IterN<-100 # Number of iterations
NatSims<-{} # Dummy objects to hold output
IntSims<-{}
```

### 2b. For loop
  * Sample, with replacement and calculate average
  
  * Store average in NatSims or IntSims

```{r}
for (i in 1:IterN){
  NatSims[i]<-mean(sample(NatSVals,length(NatSVals),replace=T))
  IntSims[i]<-mean(sample(IntSVals,length(IntSVals),replace=T))
}
```

### 2c. Calculate 95% confidence intervals

Sort from low to high
```{r}
NatSims<-sort(NatSims)
IntSims<-sort(IntSims)
```

Calculate 95%iles from simulations
```{r}
CIs<-c(sort(NatSims)[round(IterN*0.025,0)], # Native, lower 2.5%
       sort(NatSims)[round(IterN*0.975,0)], # Native, upper 97.5%
       sort(IntSims)[round(IterN*0.025,0)], # Intro, lower 2.5%
       sort(IntSims)[round(IterN*0.975,0)]) # Intro, upper 97.5%
```

### 3. Plot components

#### Combine output for plotting
```{r}
HistData<-data.frame(s=SelData$s,Native=SelData$Native)
```
*NOTE:* This creates a 'stacked' dataset for plotting

```{r}
p <- ggplot() + theme_classic()
p <- p + geom_freqpoly(data=HistData[HistData$Native=="yes",], aes(s,y=(..count..)/sum(..count..)),alpha = 0.6,colour="#1fcebd",size=2)
print(p) # native species histogram
p <- p + geom_freqpoly(data=HistData[HistData$Native=="no",], aes(s,y=(..count..)/sum(..count..)),alpha = 0.5,colour="#f53751",size=2)
print(p) # introduced species histogram
p <- p + geom_rect(aes(xmin=CIs[1],xmax=CIs[2],ymin=0,ymax=0.01),colour="white",fill="#1fcebd88")
print(p) # native species 95% CI bar
p <- p + geom_line(aes(x=mean(NatSims),y=c(0,0.01)),colour="#1d76bf",size=1)
print(p) # native species bootstrap mean
p <- p + geom_rect(aes(xmin=CIs[3],xmax=CIs[4],ymin=0,ymax=0.01),colour="white",fill="#f5375188")
print(p) # introduced species 95% CI bar
p <- p + geom_line(aes(x=mean(IntSims),y=c(0,0.01)),colour="#f53751",size=1)
print(p) # introduced species bootstrap mean
p <- p + ylab("Frequency") + scale_x_continuous(limits = c(0, 1.5))
print(p) # labels added, truncated x-axis
```

<br>

***

<br>

# 4. Custom theme
Customize various aspects such as font size, line widths, colours, etc.
```{r}
theme_custom<-function(base_size = 24, base_family = ""){
  theme_bw(base_size = base_size, base_family = base_family) %+replace% 
    theme(axis.text = element_text(size = rel(0.8),colour="black"),
          axis.ticks = element_line(colour = "black"), 
          legend.position = "none",
          panel.border = element_rect(fill = NA, colour = NA), 
          panel.grid.major = element_line(colour = NA,size = 0), 
          panel.grid.minor = element_line(colour = NA,size = 0), 
          axis.line = element_line(colour ="black")
    )
}
p + theme_custom()
```

<br>

***

<br>

# 5. Multi-graph

## Setup
Install 'gridExtra' if you haven't already
`install.packages("gridExtra")`
```{r}
library(gridExtra)
```

## grid.arrange()
#### Combine multiple plots
Prints graphs in rows, then columns, from top left to bottom right

Use ***nrow =*** and ***ncol =*** to control layout 
```{r}
grid.arrange(p,BivPlot,BarPlot,ncol=1)
grid.arrange(p,BivPlot,BarPlot,nrow=2)
```

Apply consistent formatting
```{r}
HistPlot<-p + theme_custom()
BarPlot<-BarPlot + theme_custom()
BivPlot<-BivPlot + theme_custom()
grid.arrange(HistPlot,BivPlot,BarPlot,HistPlot,nrow=2)
```

## viewport & newpage

Install 'gridExtra' if you haven't already
`install.packages("grid")`
```{r}
library(grid)
```

Control layout as a custom grid
```{r}
grid.newpage() # Open a new page on grid device
pushViewport(viewport(layout = grid.layout(3, 2))) # Create 3x2 grid layout
print(HistPlot, vp = viewport(layout.pos.row = 3, layout.pos.col = 1:2)) # Add fig in row 3 and across columns 1:2
print(BivPlot, vp = viewport(layout.pos.row = 1:2, layout.pos.col = 1)) # add fig acros rows 1:3 in column 1
print(BarPlot, vp = viewport(layout.pos.row = 1:2, layout.pos.col = 2))
```

Use viewport to add insets
```{r}
HistPlot
pushViewport(viewport(layout = grid.layout(4, 4))) # Create 4x4 grid layout (number of cells, will determine size/location of graph)
print(BivPlot, vp = viewport(layout.pos.row = 1:2, layout.pos.col = 3:4))
```

<br>

***

<br>

# 6. Reference
The comprehensive source for ggplot by Hadley Wickham:

http://link.springer.com/book/10.1007%2F978-0-387-98141-3

http://moderngraphics11.pbworks.com/f/ggplot2-Book09hWickham.pdf



