---
title: |-
  R Crash Course \
   Part 4 -- Regular Expressions in R
author: "Rob Colautti"
---


# 1. Overview

Regular Expressions

* aka 'regex' and 'regexp'

* a sort of find-and-replace for nerds

* one of the most powerful data tools I have ever learned

* requires patience and lots of **practice**


# 2. Basic Regex Functions in R

## Four Basic Regex Functions in R:

`grep()` and `grepl()` are equivalent to 'find' in your favorite word processor

* General form: `grep("find this", in.this.object)`

`sub()` and `gsub()` are equivalent to 'find and replace'

* General form: `grep("find this", "replace with this", in.this.object)`

## Three Advanced Regex Functions in R:

`regexpr()` provides more detailed info about the first match

`gregexpr()` provides more detailed results about all matches

More examples [here](https://bioinfomagician.wordpress.com/2013/11/03/regular-expression-tutorial-2-commands-in-r/)

## Start with a simple data frame of species names:

```{r}
Species<-c("petiolata", "verticillatus", "salicaria", "minor")
print(Species)
```

## `grep()` -- returns cell addresses matching query 

```{r}
grep("a",Species)
```

## `grepl()` -- returns T/F associated with 

```{r}
grepl("a",Species)
```

## `sub()` -- replaces first match (in each cell)

```{r}
sub("l","L",Species)
```

## `gsub()`  -- replaces all matches

```{r}
gsub("l","L",Species)
```


# 3. Wildcards

## `\` escape character

## `\\` in R, double-escape is usually needed (because first `\` is used to escape special characters in R, and `\` is itself a special character that needs to be escaped to pass through the function")

## `\\w`= all letters and digits (aka 'words')
```{r}
gsub("w","*","...which 1-100 words get replaced?")
gsub("\\w","*","...which 1-100 words get replaced?")
```

## `\\W` = non-word and non-number (inverse of `\\w`)
```{r}
gsub("\\W","*","...which 1-100  words get replaced?")
```

## `\\s` = spaces
```{r}
gsub("\\s","*","...which 1-100  words get replaced?")
```

## `\\t` = tab character (useful for tab-delimited data)
```{r}
gsub("\\t","*","...which 1-100  words get replaced?")
```

## `\\d` = digits
```{r}
gsub("\\d","*","...which 1-100  words get replaced?")
```

## `\\D` = non-digits
```{r}
gsub("\\D","*","...which 1-100  words get replaced?")
```

## __Two more special wildcards:__

## `\\r` = carriage return

## `\\n` = newline character

Unix/Mac files -- lines usually end with `\\n` only

Windows/DOS files -- lines usually end with `\\r\\n`


# 4. Special characters:

## `|`= 'or'

Example, look for w or e
```{r}
gsub("w|e","*","...which 1-100  words get replaced?")
```

## `.` = any character except new line
```{r}
gsub(".","*","...which 1-100  words get replaced?")
```

So how to search for a period? Use the escape character
```{r}
gsub("\\.","*","...which 1-100  words get replaced?")
```

## Use *, ? + and {} for more complicated searches

## Look at these examples carefully
```{r}
sub("\\w","*","...which 1-100 words get replaced?")
gsub("\\w","*","...which 1-100 words get replaced?")
```

## `+` = 1 or more matches
```{r}
sub("\\w+","*","...which 1-100 words get replaced?")
gsub("\\w+","*","...which 1-100 words get replaced?")
```


## `?` is a 'lazy' search (find the first occurrance)
```{r}
sub("\\w?","*","...which 1-100 words get replaced?")
gsub("\\w?","*","...which 1-100 words get replaced?")
```

## `*` is a 'greedy' search (find the last occurrence)
```{r}
sub("\\w*","*","...which 1-100 words get replaced?")
gsub("\\w*","*","...which 1-100 words get replaced?")
```

## `{n,m}` = between n to m matches
```{r}
gsub("\\w{3,4}","*","...which 1-100 words get replaced?")
```

## `{n}` = exactly n matches
```{r}
gsub("\\w{3}","*","...which 1-100 words get replaced?")
```

## `{n,}`= n or more matches
```{r}
gsub("\\w{4,}","*","...which 1-100 words get replaced?")
```


# 5. List range of options using []

## Find everything in square brackets
```{r}
gsub("[aceihw-z]","*","...which 1-100 words get replaced?")
```

## Find everything in square brackets occurring 1 or more times
```{r}
gsub("[aceihw-z]+","*","...which 1-100 words get replaced?")
```

# 6. ^Start and end of line$

## ^ Start of line

Find species starting with "s"
```{r}
grep("^s",Species)
```

### IMPORTANT: ^ Also 'negates' when used within []

Find species containing any letter other than s
```{r}
grep("[^a]",Species)
```

Replace every letter except s
```{r}
gsub("[^a]","*",Species)
```

## $ End of line

Find species ending with "a"
```{r}
grep("a$",Species)
```


# 7. Capture text with brackets `()`

Capture text using `()` and reprint using `\\1`, `\\2`, etc

Replace each word with its first letter
```{r}
gsub("(\\w)\\w+","\\1","...which 1-100 words get replaced?")
```

Pull out only the numbers and reverse their order
```{r}
gsub(".*([0-9]+)-([0-9]+).*","\\2-\\1","...which 1-100 words get replaced?")
```

Reverse first two letters of each word
```{r}
gsub("(\\w)(\\w)(\\w+)","\\2\\1\\3","...which 1-100 words get replaced?")
```

# 8. 'Scraping' data from online sources

Use `readLines` with `curl()` to 'scrape' data from the internet, then `grep()` and `gsub()` into useful structure.

```{r}
library(curl) 
Prot<-readLines(curl("http://www.rcsb.org/pdb/files/1ema.pdb"))
Prot[grep("TITLE",Prot)]
```

> HINT: You may have to use install.packages("curl") the first time you run this, if you don't already have it installed.

# EXAMPLE 1: Genbank Record

In this example, we'll work with a download of the 18S rRNA gene from *Lythrum salicaria* You  manually:

  1. Open [Lythrum 18S rRNA gene](https://www.ncbi.nlm.nih.gov/nuccore/7595475/) in a web browser
  2. Look for as a 'Send To:' button and choose 'File'
  3. Choose GenBank format and save to your computer (e.g. you current project folder for this tutorial)
  4. You should have a file called `sequence.gb` -- this is just a text file with the `.gb` extention to tell you it's a genbank layout.
  5. Import the file into R as an object using the `scan` function

```{r}
Lythrum.18S<-scan("./sequence.gb",what="character",sep="\n")
```

The 18S gene s equence for *Lythrum salicaria* (from Genbank)
```{r}
print(Lythrum.18S)
```

Notice that each line is read in as a separate cell in a vector, with sequences beginning with a number ending with 1. We can take advantage of this to extract just the sequence data

Use `.*` with `()` to delete everything before the DNA sequence
```{r}
Seq<-gsub(".*(1 [gatc])","",paste(Lythrum.18S))
paste(Seq,collapse="")
```

Use the `.*` and space with `+` to eliminate all text before the sequence :

```{r}
Seq<-gsub(".*ORIGIN +","",paste(Seq,collapse=""))
Seq
```

Elimminate spaces and the two `//` at the end

```{r}
Seq<-gsub(" |//","",Seq)
Seq
```

Capital letters look nicer, but requires a PERL qualifier `\\U` that is not standard in R
```{r}
Seq<-gsub("([actg])","\\U\\1",Seq,perl=T)
print(Seq)
```

Look for start codons?
```{r}
gsub("ATG","-->START<--",Seq)
```

Open reading frames?
```{r}
gsub("(ATG([ATGC]{3})+?(TAA|TAG|TGA)+?)","-->\\1<--",Seq)
```

Or go back, and keep non-reading frame in lower case
```{r}
Seq<-gsub(".*ORIGIN","",Lythrum.18S)
Seq<-gsub("[\n0-9 /]+","",Seq) ## Note the single \ in this case since it is inside the square brackets
gsub("(atg([atgc]{3})+?(taa|tag|tga)+?)","\\U\\1",Seq,perl=T)
```


# EXAMPLE 2: Organizing Data
Imagine you have a repeated measures design. 3 transects (A-C) and 3 positions along each transect (1-3)
```{r}
Transect<-data.frame(Species=1:20,A1=rnorm(20),A2=rnorm(20),A3=rnorm(20),B1=rnorm(20),B2=rnorm(20),B3=rnorm(20),C1=rnorm(20),C2=rnorm(20),C3=rnorm(20))
head(Transect)
```

You want to look at only transect A for the first 3 species
```{r}
Transect[1:3,grep("A",names(Transect))]
```

Or look at the first position of each transect for the first 3 species
```{r}
Transect[1:3,grep("1",names(Transect))]
```

Or rows with species IDs containing the number 2
```{r}
Transect[grep("2",Transect$Species),]
```

# PRACTICE EXERCISES

## 1. Consider a vector of email addresses scraped from the internet:

  * robert 'dot' colautti 'at' queensu 'dot' ca
  * chris.eckert[at]queensu.ca
  * lonnie.aarssen at queensu.ca

Use regular expressions to convert all email addresses to the standard format: name@queensu.ca

## 2. Create a random sequence of DNA:
```{r,results="hide"}
My.Seq<-sample(c("A","T","G","C"),1000,replace=T)
```

    * Replace T with U
    * Find all start codons (AUG) and stop codons (UAA, UAG, UGA)
    * Find all open reading frames (hint: consider each sequence beginning with AUG and ending with a stop codon; how do you know if both sequences are in the same reading frame?)
    * Count the length of bp for all open reading frames

## 3. More online examples

[http://regex.sketchengine.co.uk/extra_regexps.html](http://regex.sketchengine.co.uk/extra_regexps.html)

## 4. Regex Golf

Have fun! [LINK](https://alf.nu/RegexGolf)

