---
title: |-
  R Crash Course \
   Part 5 -- Advanced Techniques
author: "Rob Colautti"
---

# 1. Custom functions

## General form:
```{r,eval=FALSE}
functionName<-function(var1=Default1,var2=Default2){
  ## Meat and potatoes script
}
```

## Example function
User inputs two objects; the function outputs a list of functions applied to the inputs
```{r}
my.function<-function(var1=0,var2=0){
  # You can make new variables within a function
  add<-var1+var2
  subt<-var1-var2
  mult<-var1*var2
  div<-var1/var2  
  outlist<-list(input1=var1,input2=var2,addition=add,subtraction=subt,multiplication=mult,division=div)
  # So far, everything is contained within the function. 
  # Use return() to generate output
  return(outlist)
}
```
Note: Nothing output when the funciton is run. This just loads the function into memory.

### Run the function
```{r}
my.function(var1=10,var2=0.1)
my.function(var1=c(1:10),var2=c(10:1))
```

#### Protip #1:
Consider annotating long or complex script with cat() to help with troubleshooting

Note: \n = new line
```{r}
my.function<-function(var1=0,var2=0,verbose=FALSE){
  cat("\nInput variables:\nvar1 =",var1,"\nvar2 =",var2,"\n")
  cat("\nCalculating functions...\n")
  cat("\nAdding...\n")    

  add<-var1+var2
  
  cat("\nSubtracting...\n")

  subt<-var1-var2
  
  cat("\nMultiplying...\n")
  
  mult<-var1*var2
  
  cat("\nDividing...\n")
  
  div<-var1/var2  
  
  cat("\nGenerating output...\n\n")
  
  outlist<-list(input1=var1,input2=var2,addition=add,subtraction=subt,multiplication=mult,division=div)

  return(outlist)
}

## Run
my.function(var1=10,var2=0.1)
```

#### Better yet, make it an option:
```{r}
my.function<-function(var1=0,var2=0,verbose=FALSE){
  if (verbose==T){
    cat("\nInput variables:\nvar1 =",var1,"\nvar2 =",var2,"\n")
    cat("\nCalculating functions...\n")
    cat("\nAdding...\n")    
  }
  
  add<-var1+var2
  
  if (verbose==T){
    cat("\nSubtracting...\n")
  }
  
  subt<-var1-var2
  
  if (verbose==T){
    cat("\nMultiplying...\n")
  }
  
  mult<-var1*var2
  
  if (verbose==T){
    cat("\nDividing...\n")
  }
  
  div<-var1/var2  
  
  if (verbose==T){
    cat("\nGenerating output...\n")
  }
  
  outlist<-list(input1=var1,input2=var2,addition=add,subtraction=subt,multiplication=mult,division=div)

  return(outlist)
}

# Run
my.function(var1=10,var2=0.1,verbose=FALSE)
my.function(var1=10,var2=0.1,verbose=TRUE)
```


#### Protip #2:
If you have a custom function, theme, script, etc., that you use repeatedly:

  1. Save in a separate file
  
  * e.g. make new "myfunction.R" file containing just my.function
  
  2. Load using `source("PathName.FileName.R")`

  * e.g. `source("C:/Users/Colautti/Documents/RFunctions/myfunction.R")` if save

#### Protip #3:
We have already been using functions that somebody else wrote in R.

To see 'under the hood' type a function without the bracktes:

```{r}
my.function
library(ggplot2)
qplot
```

<br>

***

<br>

# 2. R Markdown

In RStudio: File->New->R Markdown
Choose html or Word (trivial to switch later)
PDF requires LaTeX

> NOTE: If you don't see an option for R Markdown, you may need to install it:

```{r,eval=FALSE}
install.packages("rmarkdown")
```


Very few elements are needed for a basic markdown file

## YAML Header 

This generated automatically when you make a new .Rmd file in RStudio. 
```{r,eval=FALSE}
---
title: R Crash Course
author: "Rob Colautti"
date: "Thursday, February 18, 2016"
output: word_document
---
```

## Basic elements:

embed R code inline using \` characters: \``embedded code`\`

Or block code using:

\`\`\`\{r\}
```{r,eval=FALSE}
block code
```
\`\`\`

For formatting text, see:

http://rmarkdown.rstudio.com/authoring_basics.html

A quick cheat-sheet is nice to have printed out for reference:

https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf

## Updates:

Update to latest version of rmarkdown
```{r, eval=FALSE}
devtools::install_github("rstudio/rmarkdown")
```

## Options for header
```{r eval=FALSE}
output:
  html_document: # Add options for html output
    toc: true # Add table of contents (TOC)
    number_sections: true # Add section numbers
    toc_float: # Have TOC floating at the side, rather than at the beginning
      collapsed: false # Expand subsections
```

## Content as tabs
```{r eval=FALSE}
## Quarterly Results {.tabset}

### By Product

(tab content)

### By Region

(tab content)
```

Looks like this:

## Quarterly Results {.tabset}

### By Product

(tab content)

### By Region

(tab content)


## Equations

Insert equations using LaTeX. Here is a handy [cheat sheet](./latexsheet.pdf)

For example:
```{r eval=FALSE}
$$\idotsint_V \mu(u_1,\dots,u_k) \,du_1 \dots du_k$$ 
$$\sum_{n=1}^{\infty} 2^{-n} = 1$$ 
```

$$\idotsint_V \mu(u_1,\dots,u_k) \,du_1 \dots du_k$$ 

$$\sum_{n=1}^{\infty} 2^{-n} = 1$$ 

Or inline:

```{r eval=FALSE}
Product $\prod_{i=a}^{b} f(i)$ inside text
```

Product $\prod_{i=a}^{b} f(i)$ inline with text

<br>

***

<br>

# 3. Custom packages

The comprehensive source on custom packages:

http://r-pkgs.had.co.nz/

(Hadley Wickham again)

### Install latest version of R
#### Important:
Quit R Studio before installing latest version of R

Download here:

https://cran.r-project.org/bin/windows/base/

### Install packages
```{r, eval=FALSE}
install.packages(c("devtools", "roxygen2", "testthat", "knitr"))
```


### Check installation
Make sure you have installed the preview version of R Studio

The following should return 'TRUE'
```{r, eval=F}
install.packages("rstudioapi")
rstudioapi::isAvailable("0.99.149")
```
If you see 'FALSE', try re-installing the preview version

http://www.rstudio.com/products/rstudio/download/preview/

### Keep devtools up to date
```{r, eval=F}
install.packages("curl")
devtools::install_github("hadley/devtools")
```

### Check library
```{r, eval=F}
library(devtools)
has_devel()
```

### Use RStudio
RStudio 'projects' make it easy to make a new package

  1. File-->New Project

  2. --> New Directory
  
  3. --> R Package
  
  4. --> Add files (data, scripts, function, theme, etc.)
  
  5. Give it a name: e.g. 'myFirstPackage'
  
### Modify files:
See 'Files' tab of R Studio

  1. myFirstPackage.proj
  
  2. DESCRIPTION
  
  3. Vignettes folder -- contains custom R markdown files
      
  4. Data in main folder; no need to setwd()
  
  
### Basic package complete!

#### Before going public:

  1. Create a github repository
  
  2. Flesh out documentation and meta-data
  
  3. Error checking
  
  4. Other tips:  http://r-pkgs.had.co.nz/release.html







